---
title: "Survival After Myocardial Infarction: A Reproducible Re-Analysis"
subtitle: "How an inverted event indicator manufactured significant findings"
author: "Tyler Tschanz"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.align = "center", fig.width = 7, fig.height = 4.5)

root <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."))
source(file.path(root, "R", "echo_survival.R"))
library(knitr)

raw      <- load_echocardiogram(file.path(root, "data", "echocardiogram_raw.csv"))
prep     <- prepare_survival_data(raw)
dat      <- prep$cohort
model_df <- prep$model
plog     <- prep$log
co       <- summarise_cohort(dat)

km_pe  <- fit_km(dat, "pericardial_effusion")
lr_pe  <- logrank(dat, "pericardial_effusion")
kc_pe  <- km_counts(dat, "pericardial_effusion")
km_wmi <- fit_km(dat, "wmi_group")
lr_wmi <- logrank(dat, "wmi_group")
kc_wmi <- km_counts(dat, "wmi_group")

kc_line <- function(kc) sprintf(
  "n = %d (%d deaths, %d censored; %d excluded for a missing grouping value)",
  kc$n_used, kc$events, kc$censored, kc$n_missing_group)

cox <- fit_cox(model_df)
hr  <- tidy_hazard_ratios(cox)
ph  <- check_ph(cox)
cstat <- summary(cox)$concordance[["C"]]
lrt   <- summary(cox)$logtest

fmt_p <- function(p) ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))
```

## Summary

This project re-analyses the public UCI Echocardiogram cohort
(`r co$n` patients followed after myocardial infarction) using
time-to-event methods. Its purpose is methodological. A prior version of
this analysis defined the survival event as *being alive* rather than
*death*, an inverted indicator that fabricated strong, statistically
significant associations. With the event coded correctly, **those
associations disappear**: no echocardiographic variable significantly
predicts survival in this small cohort. The value here is the
demonstration of correct survival methodology and the discipline to
report a null result rather than a convenient one.

## The methodological error

In this dataset the status field `still_alive` is `1` when the patient
was alive at last contact (a **right-censored** observation) and `0` when
the patient **died** (the event of interest). The survival object must
therefore be built with the event equal to `still_alive == 0`.

The prior analysis used:

```r
Surv(time = survival, event = still_alive)   # WRONG: alive coded as the event
```

This treats survivors as having experienced the event and decedents as
censored. The resulting Kaplan-Meier curves, log-rank tests, and Cox
model all estimate the hazard of *remaining alive*, which is not a
meaningful quantity. Because the scrambled event/censoring structure
remained correlated with the covariates, the broken model still produced
small p-values, which were then interpreted as clinical findings. The
correct specification is:

```r
Surv(time = survival_months, event = (still_alive == 0))
```

Every result below uses the correct specification.

## Data and provenance

The UCI Echocardiogram dataset records survival time, vital status, and
echocardiographic measurements for patients after a heart attack. It is
public, de-identified, and distributed under CC BY 4.0
(DOI 10.24432/C5QW24); the `name` field is a constant placeholder. Four
source columns are excluded from analysis: the raw wall-motion score
(superseded by the index), an undocumented derivative field (`mult`),
the documented-meaningless `group` field, and `alive_at_1yr`. The last
is a **post-outcome derived label**: it is determined by survival time
and status at 12 months, so using it to predict survival would be target
leakage.

The source file contains one structurally malformed record (line 50: a
spurious leading comma yields 14 fields instead of 13). A lenient
`read.csv` would silently wrap that line across two rows and corrupt the
values, the same class of quiet data defect this project exists to
expose. Ingestion therefore validates the field count of every line and
**excludes** the malformed record explicitly rather than guessing at it.

```{r prep-table}
data.frame(
  Step = c("Source records",
           sprintf("Excluded: malformed (line %s)",
                   paste(plog$malformed_lines, collapse = ", ")),
           "Well-formed records",
           "Removed: missing or invalid time/status",
           "Analysis cohort (Kaplan-Meier, log-rank)",
           "Removed: missing model covariate",
           "Complete-case cohort (Cox model)"),
  n = c(plog$n_source, plog$n_malformed, plog$n_wellformed,
        plog$n_dropped_outcome, plog$n_valid_outcome,
        plog$n_dropped_covariate, plog$n_complete_case)
) |> kable(caption = "Cohort derivation, from raw source file to model.")
```

## Cohort

```{r cohort-table}
data.frame(
  Measure = c("Patients", "Deaths (events)", "Censored (alive)",
              "Median survival (Kaplan-Meier)", "Median follow-up",
              "Age at infarction, mean (SD)"),
  Value = c(
    co$n,
    sprintf("%d (%.1f%%)", co$events, 100 * co$events / co$n),
    co$censored,
    ifelse(is.na(co$median_survival), "not reached",
           sprintf("%.0f months", co$median_survival)),
    ifelse(is.na(co$median_followup),
           sprintf("not reached (max %.0f months)", co$max_followup),
           sprintf("%.1f months", co$median_followup)),
    sprintf("%.1f (%.1f)", co$age_mean, co$age_sd)
  )
) |> kable(caption = "Cohort and censoring summary.")
```

This is a high-mortality cohort: `r co$events` of `r co$n` patients
(`r sprintf("%.0f%%", 100 * co$events / co$n)`) died during follow-up.
Median follow-up is reported by the reverse Kaplan-Meier estimator, the
correct estimator for follow-up time; it is not reached because fewer
than half the patients are censored.

## Kaplan-Meier analysis

### By pericardial effusion

`r kc_line(kc_pe)`.

```{r km-pe}
km_plot(km_pe, "Survival by pericardial effusion", "Pericardial effusion")
risk_table(km_pe, times = c(0, 12, 24, 36, 48)) |>
  kable(caption = "Number at risk.")
```

The curves overlap and there is no statistically significant difference
(log-rank &chi;&sup2; = `r sprintf("%.2f", lr_pe$chisq)`,
p = `r fmt_p(lr_pe$p_value)`).

### By wall-motion index

A pre-specified, clinically interpretable split is used for display
only: an index of 1.0 means every visualised segment was normal, and a
value above 1.0 indicates abnormal wall motion. The Cox model below keeps
the index continuous and applies no cut-point.

`r kc_line(kc_wmi)`.

```{r km-wmi}
km_plot(km_wmi, "Survival by wall-motion index", "Wall-motion index")
risk_table(km_wmi, times = c(0, 12, 24, 36, 48)) |>
  kable(caption = "Number at risk.")
```

Again there is no statistically significant difference (log-rank
&chi;&sup2; = `r sprintf("%.2f", lr_wmi$chisq)`,
p = `r fmt_p(lr_wmi$p_value)`).

## Cox proportional-hazards model

A multivariable Cox model was fitted on the `r plog$n_complete_case`
complete cases (`r summary(cox)$nevent` deaths) with age, pericardial
effusion, fractional shortening, E-point septal separation, left
ventricular diastolic dimension, and the continuous wall-motion index.

```{r cox-table}
data.frame(
  Term = hr$term,
  `Hazard ratio (95% CI)` = sprintf("%.2f (%.2f-%.2f)",
                                    hr$hazard_ratio, hr$ci_low, hr$ci_high),
  p = vapply(hr$p_value, fmt_p, character(1)),
  check.names = FALSE
) |> kable(caption = "Multivariable Cox hazard ratios.")
```

No covariate reaches significance. The model's discrimination is
effectively chance (Harrell's C = `r sprintf("%.3f", cstat)`) and it does
not improve on the null model (likelihood-ratio test
p = `r fmt_p(lrt[["pvalue"]])`). The wide confidence intervals are
consistent with a small sample and few events per variable.

### Proportional-hazards assumption

```{r ph-table}
data.frame(
  Term = ph$table$term,
  chisq = round(ph$table$chisq, 2),
  df = ph$table$df,
  p = vapply(ph$table$p, fmt_p, character(1)),
  check.names = FALSE
) |> kable(caption = "Scaled Schoenfeld residual test (cox.zph).")
```

The global test (p = `r fmt_p(ph$global_p)`) and every per-term test are
non-significant, so there is **no evidence of a proportional-hazards
violation** (which is not the same as proving the assumption holds). It
is also largely moot here: there is no effect to model.

## Interpretation

With the event indicator corrected, the previously reported "significant
predictors of survival" vanish entirely. The honest conclusion is the
absence of a finding: in this small, single-source cohort there is **no
statistically significant evidence** that these echocardiographic
measurements stratify post-infarction survival, and the multivariable
Cox model shows no discriminative ability. This is *absence of evidence*,
not evidence of absence: the cohort is small and underpowered, so a true
moderate effect could easily go undetected. What can be said firmly is
the negative methodological result, that the earlier significant findings
were artifacts of the inverted event coding rather than clinical signal.

This is the intended result. The objective was not to find predictors;
it was to apply survival methodology correctly and report what the data
actually support.

## Limitations

- **Time origin and predictor timing.** The time scale is months since
  myocardial infarction, but the dataset does not document *when* each
  echocardiogram was performed relative to the infarction. If imaging
  occurred at variable, post-infarction times, patients had to survive
  to be measured, which introduces survivor / immortal-time bias. These
  results must therefore be read as a descriptive, retrospective
  association exercise, **not** as causal or prospective clinical
  prediction.
- Small cohort (`r co$n` patients; `r plog$n_complete_case` complete
  cases) with few events per covariate, so confidence intervals are wide
  and the study is underpowered for modest effects (absence of evidence,
  not evidence of absence).
- Single-source observational data with documented quality issues
  (one malformed source record, placeholder/meaningless columns, missing
  values).
- The wall-motion dichotomy is for visualisation only and is not used
  for inference.
- No external validation cohort; findings describe this dataset only.

## Reproducibility

The analysis logic lives in `R/echo_survival.R`. Both the command-line
runner (`R/run_analysis.R`) and this report source that single file, so
the script and the report cannot disagree. `tests/test_pipeline.R`
asserts the correct event coding and the expected row counts at every
filtering step, so a regression in either fails loudly. Re-run end to
end with:

```sh
Rscript R/run_analysis.R          # console summary + figures
Rscript tests/test_pipeline.R     # assertions (non-zero exit on failure)
Rscript -e 'rmarkdown::render("report/survival_analysis.Rmd")'
```

```{r session, echo=TRUE}
sessionInfo()
```

## References

Salzberg, S. (1988). *Echocardiogram* [Data set]. UCI Machine Learning
Repository. CC BY 4.0. DOI 10.24432/C5QW24.
https://archive.ics.uci.edu/dataset/38/echocardiogram

Therneau, T. (2024). *A Package for Survival Analysis in R*. R package
version 3.x. https://CRAN.R-project.org/package=survival
